\documentclass[11pt,letterpaper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{geometry}
\geometry{
  top=1in,
  bottom=1in,
  left=0.75in,
  right=0.75in
}

\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cite}

\title{\LARGE \bf Astrum Verum: Composition-Episodic Cognitive Memory for LLM Agents via Vector Symbolic Architectures}

\author{
Vitaliy Fedotov \\
Independent Researcher \\
\texttt{vitaliy.fedotov.pro@gmail.com}
}

\date{}

\begin{document}

\maketitle

\begin{abstract}
Large Language Models (LLMs) equipped with standard Retrieval-Augmented Generation (RAG) suffer from structural blindness: they cannot distinguish between role-swapped propositions (e.g., "Alice trusts Bob" versus "Bob trusts Alice") due to the commutative nature of continuous vector search. We introduce a composition-episodic cognitive memory architecture based on Vector Symbolic Architectures (VSA). By decoupling semantic embedding from structural role-binding, the system preserves exact structural relationships and sequential episodic order without sacrificing continuous generalization. We benchmark the architecture on a synthetically generated corpus of 16,000 facts using real LLM embeddings ($D=384$ projected to hypervectors of $D=10,000$). The system demonstrates extreme robustness: structural recall degrades gracefully from 93.3\% at 1,000 facts to 74.7\% at 16,000 facts, while episodic order recall remains near 96.5\% when constrained to a bounded working window ($W=150$). Furthermore, SimHash grounding preserves transformer geometry with a Pearson correlation of $0.992$, proving that dense semantic neighborhoods can be perfectly mapped onto orthogonal high-dimensional associative spaces.
\end{abstract}

\section{Introduction}
\label{sec:intro}

The integration of external memory into Large Language Models (LLMs) via Retrieval-Augmented Generation (RAG) \cite{lewis2020rag} has become the standard method to overcome token limits and provide agents with external knowledge. However, flat continuous vector search---the engine powering standard RAG---is inherently blind to structure. Because sentence embeddings rely on summing or averaging attention over token distributions, propositions that share identical lexemes but differ in relational structure often collapse into identical dense neighborhoods. For example, "Alice killed Bob" and "Bob killed Alice" produce near-identical cosine similarities, leading LLM agents to hallucinate roles when queried \cite{kanerva2009hyperdimensional}.

We propose \textbf{Astrum Verum}, a cognitive memory architecture that solves structural blindness by applying Vector Symbolic Architectures (VSA) \cite{gayler2004vsa, kanerva2009hyperdimensional} on top of modern transformer embeddings. Rather than using the LLM as an implicit knowledge base, we utilize it strictly as a linguistic parser to extract semantic components (subject, relation, object). These components are grounded into $10,000$-dimensional bipolar hypervectors using SimHash projection, and then algebraically bound and bundled using Multiply-Add-Permute (MAP) operations.

Our approach revives Sparse Distributed Memory (SDM) principles \cite{kanerva1988sdm} in a modern context. By separating semantic grounding from structural encoding, we enable \textit{associative} recall (by meaning) alongside \textit{structural} recall (by role), creating a memory that reasons over symbols rather than just retrieving text chunks.

\section{Related Work}
\label{sec:related}

\textbf{Vector Symbolic Architectures (VSA)} and \textbf{Holographic Reduced Representations (HRR)} \cite{plate1995holographic} provide a mathematical framework for representing compositional structure in distributed representations, extending early work on Tensor Product Variable Binding \cite{smolensky1990tensor}. Kanerva's Sparse Distributed Memory (SDM) \cite{kanerva1988sdm} established the theoretical foundation for high-dimensional associative memory, where random vectors are nearly orthogonal. Recent advances have utilized VSA for binding structures in binary sparse distributed codes \cite{rachkovskij2001representation} and emerging hardware \cite{kleyko2022vector}.

Simultaneously, the deep learning community has demonstrated that modern attention mechanisms \cite{vaswani2017attention} can be mathematically understood as approximations of SDM \cite{bricken2021attention}, and that Hopfield Networks function similarly to transformer attention \cite{ramsauer2020hopfield}. Despite this theoretical convergence, VSA and HRR have rarely been successfully bridged with the dense, continuous embeddings generated by modern transformers to solve structural blindness. Our work establishes this bridge.

\section{Methodology}
\label{sec:methodology}

\subsection{SimHash Grounding and Semantic Fidelity}
To bridge the gap between continuous dense embeddings (e.g., from Sentence-Transformers) and bipolar VSA hypervectors, we use SimHash projection \cite{charikar2002similarity}. A dense vector $v \in \mathbb{R}^d$ is multiplied by a random projection matrix $P \in \mathbb{R}^{d \times D}$. The projection matrix entries are i.i.d. Gaussian $P_{ij} \sim \mathcal{N}(0,1)$. The resulting hypervector is:
\begin{equation}
    H = \mathrm{sgn}(v P), \quad H \in \{-1, 1\}^D
\end{equation}
This ensures that semantically similar tokens remain geometrically proximate in the high-dimensional space, effectively mapping continuous semantic neighborhoods onto an orthogonal boolean hypercube.

\subsection{Vector Symbolic Architecture (VSA)}
We employ bipolar hypervectors $\{-1, 1\}^D$ ($D=10,000$) and the MAP algebra:
\begin{itemize}
    \item \textbf{Binding ($\otimes$):} Element-wise multiplication (XOR in binary) is used to bind roles to fillers (e.g., $Role_{subj} \otimes Filler_{Alice}$).
    \item \textbf{Bundling ($\oplus$):} Element-wise addition followed by a majority sign operation $\text{sgn}()$ combines multiple elements.
    \item \textbf{Permutation ($\Pi$):} Cyclic shifts encode episodic sequences, mapping temporal order into the geometric space.
\end{itemize}

A fact is encoded as a bundled hypervector of bound roles:
\begin{equation}
    Fact = \mathrm{sgn}(R_{subj} \otimes S + R_{rel} \otimes R + R_{obj} \otimes O)
\end{equation}

\subsection{Item Memory and Match Quality}
A known challenge in VSA is capacity collapse: bundling thousands of facts into a single global superposition vector rapidly causes the noise to overwhelm the signal. To circumvent this, each fact is stored as an independent hypervector in an \textit{Item Memory}. Retrieval is performed via cosine similarity between a structural probe and all stored fact vectors. This decouples system capacity from VSA superposition limits.

However, searching across an Item Memory introduces the problem of \textit{Decoded Clarity Hallucinations}. When unbinding a retrieved fact to find an answer, the expected cosine similarity for a true signal is $0.5$ ($50\sigma$ above the noise floor of $\sigma=0.01$). But across a large dataset, a non-relevant fact might randomly produce a noisy, unbound vector that happens to align closely with a dictionary concept (e.g., score = $0.502$), outscoring the true fact (score = $0.493$).

To prevent these decoding hallucinations, we define \textbf{Match Quality} ($M$) as the product of global relevance and local clarity:
\begin{equation}
    M = \text{cosine}(\text{probe}, \text{fact}) \times \text{cosine}(\text{unbound}, \text{concept})
\end{equation}
Retrieval succeeds only if the top candidate exceeds dual thresholds: $\tau_1 = 0.2$ for fact relevance and $\tau_2 = 0.35$ for decode clarity. If breached, the memory returns $\emptyset$. This formulation enables an \textbf{Early Exit} strategy: if the first retrieved candidate passes the thresholds, search terminates immediately, yielding a roughly $5\times$ speedup during retrieval.

\subsection{Bounded Working Memory}
Episodic sequences are stored by applying a cyclic permutation to the cumulative bundle for each new item. To prevent vector saturation, we maintain a sliding window of length $W = 150$: only the last $W$ items are kept in the active episodic vector. Older items are transferred to the long-term Item Memory and are not used for order recall, effectively simulating the transition from working to long-term memory.

\section{Experiments and Results}
\label{sec:experiments}

We evaluated the architecture on a synthetic but semantically diverse corpus consisting of 16,000 facts generated from real vocabulary (581 unique texts, 502 entities, 80 relations). Embeddings were generated using \texttt{paraphrase-multilingual-MiniLM-L12-v2}. All tests were run over 30 independent seeds to guarantee statistical significance, with parameters $D=10,000$ and $d=384$.

\subsection{E1: Capacity and Structural Recall}
Table \ref{tab:capacity} shows the memory's capacity to reliably recall the correct object given a (subject, relation) cue. The memory demonstrates a graceful degradation, sustaining an accuracy of 74.7\% even when storing 16,000 independent facts.

For comparison, a standard RAG system (Chroma with cosine search) was tested on the same set of role-swapped queries (e.g., `(subject, relation, ?)` vs `(?, relation, object)`). At $N=1,000$ facts, RAG achieved 61.3\% accuracy (against a random baseline of 33.3\%), while our system achieved 93.3\%.

\begin{table}[h]
\centering
\caption{Facts structural recall vs Capacity ($N$)}
\label{tab:capacity}
\begin{tabular}{rccc}
\toprule
$N$ (facts) & Mean Acc & Std Dev & Min Acc \\
\midrule
1,000  & 0.9333 & 0.0167 & 0.8900 \\
2,000  & 0.9225 & 0.0165 & 0.8900 \\
4,000  & 0.8825 & 0.0221 & 0.8250 \\
8,000  & 0.8420 & 0.0226 & 0.8050 \\
16,000 & 0.7468 & 0.0321 & 0.6950 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{E2: Episodic Order Recall}
We tested the ability to recover ordered sequences of events (episodes) using permutation binding. As shown in Table \ref{tab:episode}, tracking a bounded working memory window ($W=150$) prevents saturation and allows infinite sequence tracking with stable 96.5\% accuracy, compared to full-sequence tracking which catastrophically degrades past length 500.

\begin{table}[h]
\centering
\caption{Episodic order-recall accuracy (Window $W=150$)}
\label{tab:episode}
\begin{tabular}{rccc}
\toprule
Full Length ($L$) & Mean Acc & Std Dev & Min Acc \\
\midrule
50 & 1.0000 & 0.0000 & 1.0000 \\
100 & 0.9900 & 0.0000 & 0.9900 \\
200 & 0.9780 & 0.0031 & 0.9733 \\
500 & 0.9649 & 0.0035 & 0.9533 \\
1,000 & 0.9649 & 0.0035 & 0.9533 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{E3: SimHash Grounding Fidelity}
To prove that our hypervectors maintain the properties of the original transformer space, we measured the correlation between cosine similarities in $\mathbb{R}^{384}$ and hypervector space. Over 30 seeds, the Pearson correlation was $0.9921 \pm 0.0002$ and Spearman correlation was $0.9907 \pm 0.0002$, confirming exceptional semantic fidelity.

\section{Discussion and Conclusion}
\label{sec:conclusion}
By combining transformer embeddings with hyperdimensional algebra, Astrum Verum achieves structural cognitive memory without the combinatorial explosion of graph databases or the hallucinations of continuous RAG. 

The introduction of the Match Quality metric solves the problem of decoded clarity hallucinations, ensuring that the system returns strict boolean states ($\emptyset$ or a grounded concept) rather than forced approximations. Furthermore, utilizing an Item Memory array bypasses the traditional superposition limits of VSA. Future work will explore the integration of this cognitive memory into end-to-end LLM agent loops, hardware acceleration \cite{kleyko2022vector}, and utilizing modern VSA frameworks \cite{heddes2022torchhd} to optimize the real-time extraction of symbolic triplets from conversational context.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
